---
title: "RutiSeq-nf BBDD Report"
date: "`r Sys.Date()`"
format: 
  html:
    fig-width: 8
    fig-height: 4
    code-fold: true

params:
  projectDir: "."
  summary: "sequencing_summary.csv" # Path to the summary
  who_resistance: "bbdd/results/main/who_resistance_summary.csv"  # Path to the resistance
  clusters: "processed_clusters.tsv" # clusters lists
  output_dir: "."                         # Output directory for results
  functions_dir: "."                       # Function dataframe for all the comples scripts
---

## Database summary

The RutiSeq dataset consists of `r nrow(summary_df)` genomes, spanning `r n_lineages$n` main lineages and `r n_sublineages$n` sublineages, the most abundant of which is `r top_lineages[1, "lineage"]` (`r top_lineages[1, "count"]`), followed by `r top_lineages[2, "lineage"]` (`r top_lineages[2, "count"]`), then `r top_lineages[3, "lineage"]` (`r top_lineages[3, "count"]`).

```{r}
#| label: fig-database_summary
#| fig-cap: "Summary of taxonomic classification of MTBc genomes within the RutiSeq BBDD. (a) Genomic taxonomy by month-year of culture isolation. (b) Distribution of sublineages grouped by lineage. (c) Proportion of lineages over time."
#| warning: false
library(tidyverse)
# Load data
summary_df <- read.delim(file.path(params$projectDir, params$summary), 
                        header = TRUE, check.names = FALSE, sep = ",")

# Calculate number of lineages and sublineages
n_lineages <- summary_df |> select(lineages) |> distinct() |> count()
n_sublineages <- summary_df |> select(`Sub-lineages`) |> distinct() |> count()

# Identify the most abundant lineages
top_lineages <- summary_df |> 
  count(lineages, name = "count") |> 
  arrange(desc(count)) |> 
  head(3)
```

### Drug-resistance profiles
MTBc drug resistance for all genomes are characterised using TB-Profiler utilising the WHO database (2023). The most commond type of DR was... (@fig-dr_type_summary)

```{r}
#| label: fig-dr_type_summary
#| fig-cap: ""
#| warning: false
who_df <- read.delim(paste0(params$projectDir,params$summary,sep=""), 
                        header=TRUE, check.names=FALSE, sep = ",")


```

### Infection clusters
MTBc drug resistance for all genomes are characterised using TB-Profiler and the WHO database (@fig-infections_clusters)

```{r}
#| label: fig-infections_clusters
#| fig-cap: ""
#| warning: false
clusters_df <- read.delim(paste0(params$projectDir,params$summary,sep=""), 
                        header=TRUE, check.names=FALSE, sep = "\t")


```
