---
title: "Interactive Phylogeny"
date: today
format: 
  html:
    self-contained: true
    theme: cerulean
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
params:
  runID: "default_run"
  lineage: "default_lineage"
  RData: "default_file.Rdata"
---
## Introduction

Below are interactive maximum-likelihood phylogenies rooted on MTBC_anc ([Comas *et al*., 2013](https://pubmed.ncbi.nlm.nih.gov/23995134/)), with tips colored according to cluster assignments at specified SNP distance thresholds. An interactive table is also included, listing all identified clusters and their corresponding cluster IDs to assist in locating specific isolates.

Use this HTML report to visually inspect clusters—identify non-monophyletic groups or any potentially spurious clusters that may warrant further investigation or subdivision.

> ℹ️ Tips for exploring the phylogeny:
> - Use zooming and panning to navigate large trees.
> - The legend lets you show/hide clusters or tip labels.
> - The `Aa trace` option toggles genome/cases identifiers for readability or anonymization.
> - Double-click a clusterID in the legend to highlight only that cluster.
> - Cross-reference highlighted clusters with the interactive table below to inspect topology.

```{r, load_packages}
#| eval: true
#| echo: false
#| warning: false
packages <- c(
    "ggtree", "ape", "dplyr", "ggplot2", "plotly", "DT", "MonoPhy"
)

# Identify missing packages
missing_pkgs <- packages[!packages %in% installed.packages()[, "Package"]]

# Install missing packages
if (length(missing_pkgs) > 0) {
    install.packages(missing_pkgs, dependencies = TRUE)
}

# Load all packages
invisible(lapply(packages, library, character.only = TRUE))

set.seed(1234)
```


```{r, import_data}
#| eval: true
#| echo: false
#| warning: false
# Read environment variables
runID             <- params$runID
Rdata_file        <- params$RData
lineage_selection <- params$lineage
lineage           <- lineage_selection

#load(Rdata_file)  # loads tree_rooted and possibly other objects
load(Rdata_file)

# Build base tree with labels
i_tree <- ggtree(tree_rooted)

# Join metadata - update this if lineage-specific metadata needed
metat <- i_tree$data |>
    dplyr::inner_join(tree.clusters, by = c("label" = "SampleID"))

# Function to load data per lineage and plot data
interactive_tree <- function(threshold) {

    p <- i_tree +
    geom_point(data = metat |> filter(!is.na(.data[[threshold]])),
                aes(x = x, y = y, 
                    colour = .data[[threshold]], 
                    label = label)) +
    ggtitle(paste('SNP threshold:', threshold))

    plotly::ggplotly(p, height = 1000, width = 700) |>
        plotly::add_text(data = metat$data[metat$data$isTip, ],  x = ~x,  y = ~y,  
                        text = ~label,
                        textposition = 'right'
                        )

}
```

```{r, col_pal}
#| eval: true
#| echo: false
#| warning: false

# Extract and sort t= column names numerically
t_cols <- names(metat)[grepl("^t.", names(metat))]
t_cols_sorted <- t_cols[order(as.numeric(gsub("t.", "", t_cols)))]
t_cols_pretty <- gsub("^t\\.", "t=", t_cols_sorted)

tmp.dat <- i_tree$data
total_genomes <- tmp.dat |> select(label) |> unique() |> nrow()
```

## Summary

| **Category**                | Value                                      |
|-----------------------------|--------------------------------------------|
| **Run ID**                  | `r runID`                                  |
| **Lineage **                | `r lineage`                                |
| **Total genomes**           | `r total_genomes`                          |
| **SNP distance thresholds** | `r paste(t_cols_pretty, collapse = ", ")`  |

## Interactive phylogeny

::: {.panel-tabset}

### t=5
```{r, tree_t5}
#| eval: false
#| echo: true
#| warning: false
interactive_tree(threshold = "t.5")
```

### t=10
```{r, tree_t10}
#| eval: false
#| echo: true
#| warning: false
interactive_tree(threshold = "t.10")
```

### t=15
```{r, tree_t15}
#| eval: false
#| echo: true
#| warning: false
interactive_tree(threshold = "t.15")
```

:::

---

## Clusters

Use this interactive table to identify the clusterIDs for a particular isolate of interest. Empty cells indicate that the genome was a singleton at that clustering threshold.


## Cluster Table

```{r, dt_table}
# Reorder the full data frame
table <- metat %>%
  select(SampleID = label, lineage, all_of(t_cols_sorted)) |>
  filter(SampleID != "H37Rv") |>
  filter(SampleID != "MTB_anc")

library(DT)

# Example data frame (replace with your actual cluster data)
# Render interactive table
datatable(table, options = list(pageLength = 10, autoWidth = TRUE), filter = 'top')
```

---

## Monophylogeny test

Monophylogeny of the clusters are tested using the R pacakge `MonoPhy` (https://github.com/oschwery/MonoPhy). This will generate a summary of each genome's monophyly-status. This only applies to genomes that are in clusters.

```{r, monophyly_test}
#| eval: true
#| echo: true
#| warning: false
# create f for monophyly test (sampleID, decreasing SNP threshold clsuterIDs)
## eg: sampleID t_15  t_10  t_5
monophyl_df <- tree.clusters |> select(label = SampleID, all_of(t_cols_sorted))

# prune tree to ensure only tree in meta are in tree
# Get the list of sample labels that should remain
tips_to_keep <- monophyl_df$label

# Get all tips currently in the tree
all_tips <- tree_rooted$tip.label

# Identify tips NOT in your metadata
tips_to_drop <- setdiff(all_tips, tips_to_keep)

# Prune the tree
tree_pruned <- drop.tip(tree_rooted, tips_to_drop)

# Optional: confirm
cat("Original tips:", length(all_tips), "\n")
cat("Kept tips:", length(tree_pruned$tip.label), "\n")
cat("Dropped tips:", length(tips_to_drop), "\n")

# Run AssessMonophyly
solution1 <- AssessMonophyly(tree_pruned, monophyl_df)

# Extract TipStates tables from AssessMonophyly result
tip_status_list <- lapply(names(solution1), function(tcol) {

  df <- solution1[[tcol]]$TipStates |> 
    dplyr::select(Tip, Status) |> 
    dplyr::rename(label = Tip, !!tcol := Status)
})

# Join all t.X status columns by 'label'
status_df <- Reduce(function(x, y) dplyr::full_join(x, y, by = "label"), tip_status_list)

# Optional: ensure order matches monophyl_df
status_df <- monophyl_df |> dplyr::select(label) |> dplyr::left_join(status_df, by = "label")

# remove the unknown rows
status_df_filtered <- status_df %>%
  # ensure we’re comparing as character (avoid factor issues)
  mutate(across(all_of(t_cols_sorted), as.character)) %>%
  # keep rows where NOT all values in the int_cols_sorted are "unknown"
  filter(rowSums(across(all_of(t_cols_sorted)) == "unknown") < length(t_cols_sorted))

# Print as interative table
datatable(status_df_filtered, options = list(pageLength = 5, autoWidth = TRUE), filter = 'top')
```


