---
title: "Read Classification Report"
date: today
format: 
  html:
    self-contained: true
    theme: cerulean
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
---

```{r, load_packages}
#| eval: true
#| echo: false
#| warning: false
packages <- c(
    "tidyverse", "plotly", "DT", "UpSetR"
)

# Identify missing packages
missing_pkgs <- packages[!packages %in% installed.packages()[, "Package"]]

# Install missing packages without prompt using Bristol mirror
if (length(missing_pkgs) > 0) {
    install.packages(
        missing_pkgs,
        dependencies = TRUE,
        repos = "https://cran.ma.ic.ac.uk/"
    )
}

# Load all packages
invisible(lapply(packages, library, character.only = TRUE))

set.seed(1234)
```

```{r, test_setwd}
#| eval: false
#| echo: false
#| warning: false
setwd("/Users/poppybest/Documents/Work/Github/RutiSeq-nf/bin/quarto/")
setwd("/home/phesketh/Documents/GitHub/TBSEQ.cat-nf/bin/quarto")
```

```{r, import_df}
# sampe metadata
metat <- read.csv("samplesheet.csv", 
                  header = TRUE)
  metat$forward_base <- basename(metat$forward_path)
  metat$reverse_base <- basename(metat$reverse_path)
colnames(metat)
ska_distances <- read.delim("ska_distance.tsv")
read_stats <- read.delim("read_stats.tsv")
sylph_df <- read.delim("sylph_classification.tsv", 
                        header = TRUE, 
                        check.names = FALSE)
  colnames(sylph_df) <- sub("_R[12]\\.fastq\\.gz$", "", colnames(sylph_df))
```

## Total reads sequenced

```{r, wrangle_reads_df}
metat2 <- metat |>
  select(sampleID, forward_base, reverse_base, type) |>
  pivot_longer(
    cols = c(forward_base, reverse_base),
    names_to = "direction",
    values_to = "file"
  ) |>
  select(sampleID, dataType=type, file, direction)

read_stats <- left_join(read_stats, metat2, by = "file")
colnames(read_stats)
```

```{r, interactive_read_bp_fun}
plot_stacked_reads <- function(df = read_stats, datatype_filter) {
  
  # Filter data (allow multiple dataTypes)
  df_filt <- df %>% 
    filter(dataType %in% datatype_filter)
  
  # ggplot
  p <- ggplot(df_filt, aes(x = sampleID, y = num_seqs, fill = direction)) +
    geom_bar(stat = "identity", position = "stack") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    scale_fill_manual(values = c("#4679A3", "#A6739B")) +
    geom_hline(yintercept = 1e6, 
              color = "green", 
              linetype = "dashed",
              linewidth = 1.5)  
  ggplotly(p)
  p
}
```

::: {.panel-tabset}

### Samples
```{r, tree_t10}
#| eval: false
#| echo: true
#| warning: false
plot_stacked_reads(datatype_filter = c("sample","ENA","SRA"))
```

### Controls

```{r, import_data}
#| eval: false
#| echo: false
#| warning: false
plot_stacked_reads(datatype_filter = "control")
```

:::

## Taxonomic Classification of Reads

Reads were taxonomically classified using Sylph with the GDR220 database.

```{r, interactive_tax_bp_func}
#| eval: true
#| echo: true
#| warning: false
index <- metat |> select(identity=sampleID, type) |> distinct()

# Make ggplot compatible dataframe
sylph_long <- sylph_df %>%
  pivot_longer(
    cols = -clade_name,
    names_to = "identity",
    values_to = "perc"
  ) |>
  filter(perc != 0) |>
  separate_wider_delim(
    clade_name,
    delim = "|",
    names = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species", "Genome"),
    too_many = "debug",  # merge extra pieces into the last column
    too_few = "align_start"    # or "drop" if you don’t need debug info
  )

# Add the index information (type: sample/ENA/SRA or control)
sylph_long <- left_join(sylph_long, index)
sylph_long$perc <- as.numeric(sylph_long$perc)
```

```{r, func_interactive_tax_bp}
#| eval: true
#| echo: true
# Make function for plotting the data
interactive_tax_bp <- function(df = sylph_long,
                              type_filter = NULL,
                              tax_level = NULL) {

  # Convert tax_level (string) into a symbol for tidy evaluation
  tax_sym <- sym(tax_level)

  # Filter and select data
  plot_df <- df %>%
    filter(type %in% type_filter) %>%
    select(tax = !!tax_sym, identity, perc) %>%
    filter(!is.na(tax)) %>%
    group_by(identity, tax) %>%
    summarise(mean_perc = mean(perc), .groups = "drop") %>%
    group_by(identity) %>%
    # Add "Unknown" to make total 100 if needed
    summarise(
      tax = c(tax, if(sum(mean_perc) < 100) "NaN" else NULL),
      mean_perc = c(mean_perc, if(sum(mean_perc) < 100) 100 - sum(mean_perc) else NULL),
      .groups = "drop"
    )

    # Create a color palette for all taxa except Unclassified
    taxa <- unique(plot_df$tax)
    taxa_no_unclassified <- setdiff(taxa, "NaN")
    palette <- scales::hue_pal()(length(taxa_no_unclassified))  # default ggplot colors
    names(palette) <- taxa_no_unclassified

    # Add Unclassified as grey50
    palette["Unclassified"] <- "grey50"

    # Plot
    p <- ggplot(plot_df, aes(x = identity, y = mean_perc, fill = tax)) +
      geom_col() +
      theme_bw() +
      labs(
        x = "",
        y = "Percentage of Reads",
        fill = tax_level,
        title = paste("Taxonomic Composition at", tax_level, "Level")
      ) +
      scale_fill_manual(values = palette) +
      theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8))

    p
    #ggplotly(p)
}
```

### Samples

::: {.panel-tabset}

#### Phylum
```{r, p__samples_tax}
#| eval: true
#| echo: true
#| warning: false
interactive_tax_bp(type_filter = c("sample","ENA","SRA"),
                    tax_level = "Phylum")
```

#### Class
```{r, c__samples_tax}
#| eval: true
#| echo: true
#| warning: false
interactive_tax_bp(type_filter = c("sample","ENA","SRA"),
                    tax_level = "Class")
```

#### Order
```{r, o__samples_tax}
#| eval: true
#| echo: true
#| warning: false
interactive_tax_bp(type_filter = c("sample","ENA","SRA"),
                    tax_level = "Order")
```

#### Family
```{r, f__samples_tax}
#| eval: true
#| echo: true
#| warning: false
interactive_tax_bp(type_filter = c("sample","ENA","SRA"),
                    tax_level = "Family")
```

#### Genus
```{r, g__samples_tax}
#| eval: true
#| echo: true
#| warning: false
interactive_tax_bp(type_filter = c("sample","ENA","SRA"),
                    tax_level = "Genus")
```

#### Species
```{r, s__samples_tax}
#| eval: true
#| echo: true
#| warning: false
interactive_tax_bp(type_filter = c("sample","ENA","SRA"),
                    tax_level = "Species")
```

:::

### Controls

::: {.panel-tabset}

#### Phylum
```{r, p__controls_tax}
#| eval: true
#| echo: true
#| warning: false
interactive_tax_bp(type_filter = c("control", "controls", "CTRL"),
                    tax_level = "Phylum")
```

#### Class
```{r, c__controls_tax}
#| eval: true
#| echo: true
#| warning: false
interactive_tax_bp(type_filter = c("control", "controls", "CTRL"),
                    tax_level = "Class")
```

#### Order
```{r, o__controls_tax}
#| eval: true
#| echo: true
#| warning: false
interactive_tax_bp(type_filter = c("control", "controls", "CTRL"),
                    tax_level = "Order")
```

#### Family
```{r, f__controls_tax}
#| eval: true
#| echo: true
#| warning: false
interactive_tax_bp(type_filter = c("control", "controls", "CTRL"),
                    tax_level = "Family")
```

#### Genus
```{r, g__controls_tax}
#| eval: true
#| echo: true
#| warning: false
interactive_tax_bp(type_filter = c("control", "controls", "CTRL"),
                    tax_level = "Genus")
```

#### Species
```{r, s__controls_tax}
#| eval: true
#| echo: true
#| warning: false
interactive_tax_bp(type_filter = c("control", "controls", "CTRL"),
                    tax_level = "Species")
```

:::

## K-mer comparison between samples and controls

In this section, the results of SKA2 k-mer comparison are used to generate PCoA plots, which illustrate the differences in k-mer composition between samples. The primary goal is to identify whether any of your controls overlap substantially with your samples, which could indicate potential contamination or sample bleed-through.

```{r, ska_components}
#| eval: true
#| echo: true
#| warning: false
samples <- unique(c(ska_distances$Sample1, ska_distances$Sample2))

#Initialize empty matrix
dist_matrix <- matrix(0, nrow = length(samples), ncol = length(samples),
                      dimnames = list(samples, samples))

# Fill in distances
for(i in 1:nrow(ska_distances)) {
  s1 <- ska_distances$Sample1[i]
  s2 <- ska_distances$Sample2[i]
  d  <- ska_distances$Distance[i]
  dist_matrix[s1, s2] <- d
  dist_matrix[s2, s1] <- d   # symmetric
}

# Convert to dist object
dist_obj <- as.dist(dist_matrix)

# Run PCoA
pcoa_res <- cmdscale(dist_obj, k = 2)  # 2D

pcoa_df <- data.frame(
  Sample = rownames(pcoa_res),
  PC1 = pcoa_res[,1],
  PC2 = pcoa_res[,2]
)

pcoa_df <- left_join(pcoa_df, index, by = c("Sample" = "identity"))

# Plot
p <- ggplot(pcoa_df, 
            aes(
                x = PC1, 
                y = PC2, 
                label = Sample,
                color = type)) +
  scale_fill_manual(values = c("#2A5783", "grey50")) +
  geom_point(size = 4) +
  theme_minimal() +
  labs(title = "PCoA Plot of Kmer Distances",
      x = "PC1",
      y = "PC2")
  p
  #ggplotly(p)
```

## ⚠️ Potentially Problematic Samples and Controls

The samples and controls highlighted in the table below contain reads that are **taxonomically mixed** at the *species level*. This is especially important when analyzing reads from a **single bacterial isolate**, as it helps identify potential **cross-contamination** within the sequencing run. For example, reads from other bacterial species may have inadvertently been included in your dataset.

Controls are highlighted if they contain reads representing more than 1% of any taxonomic classification. If these reads match the species under study, this should be investigated as possible contamination. Use this table, along with the **PCoA plots above**, to assess and make informed judgments about potential contamination in your samples.

```{r, dt_table}
#| eval: true
#| echo: true
#| warning: false
# Get the identities of mixed samples
mixed_species_summary <- sylph_long %>%
  select(identity, Species, perc, type) %>%
  filter(!is.na(Species), perc > 1, perc < 95) %>%  # filter mixed species
  distinct() %>%
  group_by(identity, type) %>%
  summarise(
    `Mixed Species (>1%, <95%)` = paste0(Species, " (", perc, "%)", collapse = ", "),
    .groups = "drop"
  )

#mixed_species_summary <- left_join(mixed_species_summary, reads_summary)

# Example data frame (replace with your actual cluster data)
# Render interactive table
datatable(mixed_species_summary, options = list(pageLength = 10, autoWidth = TRUE), filter = 'top')
```
