---
title: "*Mycobacterium tuberculosis* WGS and cluster analysis"
author: RutiSeq-nf
date: today
format: 
  html:
    self-contained: true
    theme: cerulean
    toc: true
    toc-depth: 3        # Levels of headers to include
    toc-location: left  # Can be "left", "right", or "inline"
execute:
  echo: false
  warning: false
  message: false
params:
  runID: "default_run"
  analysisSummary: "default_file.analysisSummary"
  processedClusters: "default_processedClusters"
  whoResistance: "default_whoResistance"
  tbdbResistance: "default_tbdbResistance"
  samplesheet: "default_samplesheet"
  metadata: "default_metadata"
  minCov: 50
  minCovPerc: 0.98
  minReads: 500000
  iqtree_bootstraps: 1000
  iqtree_model: "GTR+G4"
  mtbseq_minbqual: 20
  mtbseq_mincovf: 4
  mtbseq_mincovr: 4
  mtbseq_minphred20: 4
  mtbseq_minfreq: 75
  mtbseq_unambig: 95
  mtbseq_window: 12
---

This interactive report provides key summary results to support the verification of your RutiSeq analysis and to flag any samples that may be unsuitable for inclusion in clustering (these can be excluded, and the analysis re-run if necessary). Samples of concern are clearly highlighted for your review. All data tables used to generate the figures are available in the `results/tables` directory of the RutiSeq output. This report is most effective for datasets containing more than 100 samples.

```{r setup, include=FALSE}
# Load required libraries
library(readr)
library(dplyr)
library(ggplot2)
library(plotly)
library(DT)
library(lubridate)
library(tidyverse)
library(ggridges)

# Read environment variables
runID             <- params$runID
#analysisSummary   <- params$analysisSummary
#processedClusters <- params$processedClusters
#whoResistance     <- params$whoResistance
#tbdbResistance    <- params$tbdbResistance
#samplesheet       <- params$samplesheet

## Analysis variables
minCov             <- params$minCov
minCovPerc         <- params$minCovPerc
minReads           <- params$minReads
iqtree_bootstraps  <- params$iqtree_bootstraps
iqtree_model       <- params$iqtree_model
mtbseq_minbqual    <- params$mtbseq_minbqual
mtbseq_mincovf     <- params$mtbseq_mincovf
mtbseq_mincovr     <- params$mtbseq_mincovr
mtbseq_minphred20  <- params$mtbseq_minphred20
mtbseq_minfreq     <- params$mtbseq_minfreq
mtbseq_unambig     <- params$mtbseq_unambig
mtbseq_window      <- params$mtbseq_window

#runID <- "test"
#minCov <- 50
#minCovPerc <- 0.95
#minReads <- 500000
#iqtree_bootstraps <- 1000
#iqtree_model <- 'GTR+G4'
#mtbseq_minbqual <- 20
#mtbseq_mincovf  <- 4
#mtbseq_mincovr <- 4
#mtbseq_minphred20 <- 4
#mtbseq_minfreq <- 75
#mtbseq_unambig <- 95
#mtbseq_window <- 12

# Load datasets
#analysisSummary_df   <- read_tsv(analysisSummary)
#processedClusters_df <- read_tsv(processedClusters)
#whoResistance_df     <- read_tsv(whoResistance)
#tbdbResistance_df    <- read_tsv(tbdbResistance)
#samplesheet_df       <- read_tsv(samplesheet)

new_samples          <- read_tsv("./test/samplesheet.csv") |> select(Sample=sampleID)
analysisSummary_df   <- read_tsv("./test/sequencingSummary.tsv")
tbdbResistance_df    <- read_tsv("./test/tbdbResistance.tsv")
whoResistance_df     <- read_tsv("./test/whoResistance.tsv")
processedClusters_df <- read_tsv("./test/clusters.tsv")
metadata_df          <- read_tsv("./test/metadata.tsv")

# Extract and sort t= column names numerically
t_cols <- names(processedClusters_df)[grepl("^t=", names(processedClusters_df))]
t_cols_sorted <- t_cols[order(as.numeric(gsub("t.", "", t_cols)))]

# Make the large summary table of the database
reduced_cluster_df <- processedClusters_df |> select(Sample=SampleID, all_of(t_cols_sorted))
analysisSummary_df_large <- left_join(analysisSummary_df, reduced_cluster_df)
```

## Database Summary

This section summarizes the entire RutiSeq-db collection along with the most recently analyzed cases. It provides an overview of the lineages of the cases, resistance profiles, and identifies any isolates that failed to meet the minimum quality metrics. These failed isolates have been excluded from the clustering analysis.

### Sequencing

```{r}
## WGS libraries read totals
library_reads <- analysisSummary_df |> select(Library, `Total Reads`)
  newest_libraries <- analysisSummary_df |> 
                        filter(Sample %in% new_samples$Sample) |> 
                        select(Library) |> unique()
newest_libraries_vec <- newest_libraries$Library

## Cases over time
analysisSummary_df <- left_join(analysisSummary_df, metadata_df, by = c("Sample" = "sampleID"))

cases_over_time <- analysisSummary_df |>
  mutate(
    date = as.Date(date),
    MonthYear = format(date, "%m-%y")
  ) |>
  select(MonthYear, lineage = `Main lineage`) |>
  count(MonthYear, lineage, name = "n") |>  # <-- FIXED LINE
  mutate(
    MonthYearDate = as.Date(paste0("01-", MonthYear), format = "%d-%m-%y")
  ) |>
  arrange(lineage, MonthYearDate) |>  # Arrange by lineage for proper cumsum
  group_by(lineage) |>                # Group by lineage for per-lineage cumsum
  mutate(cumulative_cases = cumsum(n)) |>
  ungroup()
```

In this section the raw sequence data is summarised with respect to current and past sequencing libraries. 

::: {.panel-tabset}

#### Total Reads

In this tab the total number of WGS reads are summarised by libraries. The newest libraries are colored in red.

```{r}
# Create the initial empty plot
fig <- plot_ly()

# Loop over unique libraries to add individual box traces
for (lib in unique(library_reads$Library)) {
  color <- if (lib %in% newest_libraries_vec) 'red' else 'gray'
  
  fig <- fig |>
    add_trace(
      data = filter(library_reads, Library == lib),
      x = ~Library,
      y = ~`Total Reads`,
      type = 'box',
      name = lib,
      marker = list(color = color),
      line = list(color = color),
      fillcolor = color
    )
}

# Optional layout formatting
fig <- fig |> layout(
  title = "Total Reads by Library",
  yaxis = list(title = "Total Reads"),
  xaxis = list(title = "Library")
)

fig
```

#### Cases over time

```{r}
plot_ly(
  data  = cases_over_time,
  x     = ~MonthYearDate,
  y     = ~n,
  color = ~lineage,
  type  = 'scatter',
  mode  = 'lines+markers'
) |>
  layout(
    title = "Cases Over Time",
    xaxis = list(title = "Month-Year"),
    yaxis = list(title = "Number of Cases"),
    showlegend = TRUE  # usually helpful to keep legend when coloring by lineage
  )
```

#### Cumilative cases

```{r}
plot_ly(
  data  = cases_over_time,
  x     = ~MonthYearDate,
  y     = ~cumulative_cases,
  color = ~lineage,
  type  = 'scatter',
  mode  = 'lines+markers'
) |>
  layout(
    title = "Cases Over Time (Cumilative)",
    xaxis = list(title = "Month-Year"),
    yaxis = list(title = "Number of Cases"),
    showlegend = TRUE
  )
```


:::

### Newest cases: `r runID`

```{r}
# Main lineages
new_samples_df <- analysisSummary_df |> filter(Sample %in% new_samples$Sample)

lineage_counts <- new_samples_df |> 
  count(`Main lineage`) |> 
  select(lineage = `Main lineage`, counts = n)
lineage_counts_df <- as.data.frame(lineage_counts)
rownames(lineage_counts_df) <- lineage_counts_df$lineage

# Sub lineages
sublineage_counts <- new_samples_df |> 
  count(`Main lineage`, `Sub-lineage`) |> 
  select(lineage = `Main lineage`, sublineage = `Sub-lineage`, counts = n)

# Resistance profiles (WHO)
new_samples_df <- left_join(new_samples_df, whoResistance_df, by = c("Sample" = "Sample"))
resistance_counts_df <- new_samples_df |> count(lineage = `Main lineage`, DRT = `DR type (WHO)`)

tbdbResistance_df_final <- whoResistance_df |> filter(Sample %in% new_samples$Sample) |>
                          mutate(across(
                              .cols = -c(Sample, `DR type`),
                              .fns = ~ ifelse(.x == "-", "Sensitive", "Resistant")
                            ))

# Failed QC
failed_samples_df <- new_samples_df |>
                      mutate(
                            `Unambiguous Total Bases (%)` = as.numeric(gsub(",", ".", `Unambiguous Total Bases (%)`)),
                            `Total Reads` = as.numeric(`Total Reads`),
                            `Unambiguous Coverage median` = as.numeric(`Unambiguous Coverage median`)) |> 
                      filter(
                            `Unambiguous Total Bases (%)` < minCovPerc |
                            `Total Reads` < minReads |
                            `Unambiguous Coverage median` < minCov |
                            `Infection type` == "Mixed") |>
                      select(Sample, `Total Reads`, `Unambiguous Total Bases (%)`, `Unambiguous Coverage median`,
                      `Infection type`, `Lineages (fractions)`)

# number of mixed infections
mixed_infections  <- new_samples_df |> filter(`Infection type` == "Mixed") |> nrow()
sensitive_counts  <- new_samples_df |> filter(`DR type (WHO)` == "Sensitive") |> nrow()
resistance_counts <- new_samples_df |> filter(`DR type (WHO)` != "Sensitive") |> nrow()
```

This section provides a summary of the newest cases. There were **`r nrow(new_samples_df)` new cases
**, representing **`r nrow(lineage_counts)`** distinct lineages.

**`r nrow(failed_samples_df)` genome(s) failed the minimum quality criteria** (see `Parameters`) and should be reviewed, and **`r paste(mixed_infections)` genomes were identified as mixed infection(s)** (i.e., containing multiple lineages). These are highlighted in the `Failed QC` tab.

In terms of antibiotic resistance, **`r paste(sensitive_counts)` isolates were sensitive**, while **`r paste(resistance_counts)` displayed resistance**. Resistant isolates are summarised in the `DRTypes (WHO)` and `Resistance (WHO)` tabs.

::: {.panel-tabset}

#### Main lineages
```{r}
plot_ly(lineage_counts_df, labels = ~lineage, values = ~counts, type = 'pie')
```

#### Sub-lineages
```{r}
plot_ly(
  data = sublineage_counts,
  x = ~lineage,
  y = ~counts,
  color = ~sublineage,
  type = 'bar'
) |>
  layout(
    barmode = 'stack',
    xaxis = list(title = "Lineage"),
    yaxis = list(title = "Counts"),
    legend = list(title = list(text = "Sublineage"))
  )
```

#### DRTypes (WHO)
```{r}
plot_ly(
  data = resistance_counts_df,
  x = ~DRT,
  y = ~n,
  color = ~lineage,
  type = 'bar'
) |>
  layout(
    barmode = 'stack',
    xaxis = list(title = "Drug Resistance Type (WHO)"),
    yaxis = list(title = "Counts"),
    legend = list(title = list(text = "Lineage"))
  )
```

#### Resistance (WHO)
```{r}
tbdbResistance_df_final_long <- tbdbResistance_df_final |>
    pivot_longer( cols = -c(Sample, `DR type`),
                  names_to = "Antibiotic",
                  values_to = "Resistance"
                ) |>
    mutate(
      Category = case_when(
        Antibiotic %in% c("RIF", "INH", "EMB", "PZA") ~ "1st line",
        Antibiotic %in% c("MFX", "LFX", "BDQ", "DLM", "Pa", "LZD") ~ "2nd line",
        TRUE ~ "Other"
      )
    )

hm <- ggplot(tbdbResistance_df_final_long, 
       aes(x = Antibiotic, y = Sample, fill = Resistance)) +
  geom_tile(color = "grey30", size = 0.4) +
  scale_fill_manual(values = c("S" = "grey90", "R" = "black")) +
  facet_wrap(~ Category, scales = "free_x") +
  theme_light(base_size = 10) +  # Better compatibility than theme_minimal()
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "black", color = NA),
    strip.text = element_text(face = "bold")
  ) +
  labs(
    title = "", x = "", y = "",
    fill = "Resistance"
  )

ggplotly(hm, tooltip = c("x", "y", "fill"))
```

#### Failed QC
If this table is empty, then great! All your new samples passed the quality metrics defined when you ran the RutiSeq workflow. Any failed variables will be colored in red.
```{r}
# Build datatable with conditional formatting
DT::datatable(
  failed_samples_df,
  options = list(pageLength = 10, autoWidth = TRUE),
  filter = 'top'
  ) |>
  formatStyle(
    'Total Reads',
    color = styleInterval(minReads, c('red', 'black'))
  ) |>
  formatStyle(
    'Unambiguous Total Bases (%)',
    color = styleInterval(minCovPerc, c('red', 'black'))
  ) |>
  formatStyle(
    'Unambiguous Coverage median',
    color = styleInterval(minCov, c('red', 'black'))
  ) |>
  formatStyle(
    'Infection type',
    color = styleEqual("Mixed", "red")
  )
```

#### Read taxonomy

´Sylph´ is used to taxonomically classify the WGS reads from each isolate. Highlighted here is the taxnomic ompistion of the WGS reads, highlighting potentially contaminating samples that should be investigated more closely.
```{r}

```

#### SNP freq (main)
This tab features a ridge plot displaying multiple histograms that show the number of SNP variations identified between isolates belonging to each lineage and the reference genome (H37Rv). For clarity, isolate identified as a mixed infection have been removed.

```{r}
snp_freq <- new_samples_df |> 
    filter(`Infection type` != "Mixed") |>
    select(lineage = `Main lineage`, nSNPs)
ggplot(snp_freq, 
        aes(x = nSNPs, 
            y = lineage)) +
        ggridges::geom_density_ridges(alpha = 0.6, stat = "binline") +
        ggridges::theme_ridges() + 
        theme(legend.position = "none") +
    labs(x = "No. SNPs", y = "Lineage")
```

#### SNP freq (sub)
This tab features a ridge plot displaying multiple histograms that show the number of SNP variations identified between isolates belonging to each lineage and the reference genome (H37Rv). For clarity, isolate identified as a mixed infection have been removed.

```{r}
snp_freq <- new_samples_df |> 
    filter(`Infection type` != "Mixed") |>
    select(lineage = `Sub-lineage`, nSNPs) |>
    mutate(lineage = sapply(str_split(lineage, "\\."), function(parts) {
      if (length(parts) >= 2) {
        paste(parts[1], parts[2], sep = ".")
      } else {
        parts[1]
      }
    }))

ggplot(snp_freq, 
        aes(x = nSNPs, 
            y = lineage)) +
        ggridges::geom_density_ridges(alpha = 0.6, stat = "binline") +
        ggridges::theme_ridges() + 
        theme(legend.position = "none") +
    labs(x = "No. SNPs", y = "Sub-lineage")
```

:::

### RutiSeq-db

```{r}
# Main lineages
lineage_counts <- analysisSummary_df |> 
  count(`Main lineage`) |> 
  select(lineage = `Main lineage`, counts = n)
lineage_counts_df <- as.data.frame(lineage_counts)
rownames(lineage_counts_df) <- lineage_counts_df$lineage

meta_small <- analysisSummary_df |> select(Sample, lineage = `Main lineage`, date )

# Sub lineages
sublineage_counts <- analysisSummary_df |> 
  count(`Main lineage`, `Sub-lineage`) |> 
  select(lineage = `Main lineage`, sublineage = `Sub-lineage`, counts = n)

# Resistance profiles (WHO)
analysisSummary_df <- left_join(analysisSummary_df, meta_small)
resistance_counts_df <- analysisSummary_df |> count(lineage = `Main lineage`, DRT = `DR type (WHO)`)

tbdbResistance_df_final <- whoResistance_df |> 
                          mutate(across(
                              .cols = -c(Sample, `DR type`),
                              .fns = ~ ifelse(.x == "-", "Sensitive", "Resistant")
                            ))

tbdbResistance_df_final <- left_join(tbdbResistance_df_final, meta_small)

tbdbResistance_df_final_long <- tbdbResistance_df_final |>
    pivot_longer( cols = -c(Sample, `DR type`,date, lineage),
                  names_to = "Antibiotic",
                  values_to = "Resistance"
                ) |>
    mutate(
      Category = case_when(
        Antibiotic %in% c("RIF", "INH", "EMB", "PZA") ~ "1st line",
        Antibiotic %in% c("MFX", "LFX", "BDQ", "DLM", "Pa", "LZD") ~ "2nd line",
        TRUE ~ "Other"
      )
    )

monthly_lineage_counts <- tbdbResistance_df_final_long |>
  filter(Resistance == "Resistant") |>
  mutate(month = floor_date(ymd(date), "month")) |>
  group_by(Antibiotic, month, lineage) |>
  summarise(count = n(), .groups = "drop") |>
  group_by(Antibiotic, lineage) |>
  filter(n() > 1) |>
  ungroup()

## Failed QC tables
failed_samples_df <- analysisSummary_df |>
                      mutate(
                            `Unambiguous Total Bases (%)` = as.numeric(gsub(",", ".", `Unambiguous Total Bases (%)`)),
                            `Total Reads` = as.numeric(`Total Reads`),
                            `Unambiguous Coverage median` = as.numeric(`Unambiguous Coverage median`)) |> 
                      filter(
                            `Unambiguous Total Bases (%)` < minCovPerc |
                            `Total Reads` < minReads |
                            `Unambiguous Coverage median` < minCov |
                            `Infection type` == "Mixed") |>
                      select(Sample, `Total Reads`, `Unambiguous Total Bases (%)`, `Unambiguous Coverage median`,
                      `Infection type`, `Lineages (fractions)`)

# number of mixed infections
mixed_infections  <- analysisSummary_df |> filter(`Infection type` == "Mixed") |> nrow()
sensitive_counts  <- analysisSummary_df |> filter(`DR type (WHO)` == "Sensitive") |> nrow()
resistance_counts <- analysisSummary_df |> filter(`DR type (WHO)` != "Sensitive") |> nrow()
```

This section provides a summary of the RutiSeq database, including the most recent additions. The database includes **`r nrow(analysisSummary_df)` samples**, representing **`r nrow(lineage_counts)`** distinct lineages.

**`r nrow(failed_samples_df)` genome(s) failed the minimum quality criteria** (see `Parameters`) and should be reviewed, and **`r paste(mixed_infections)` genomes were identified as mixed infection(s)** (i.e., containing multiple lineages). These are highlighted in the `Failed QC` tab.

In terms of antibiotic resistance, **`r paste(sensitive_counts)` isolates were sensitive**, while **`r paste(resistance_counts)` displayed resistance**. Resistant isolates are summarised in the `DRTypes (WHO)` and `Resistance (WHO)` tabs.

::: {.panel-tabset}

#### Main lineages

```{r}
plot_ly(lineage_counts_df, labels = ~lineage, values = ~counts, type = 'pie')
```

#### Sub-lineages

```{r}
plot_ly(
  data = sublineage_counts,
  x = ~lineage,
  y = ~counts,
  color = ~sublineage,
  type = 'bar'
) |>
  layout(
    barmode = 'stack',
    xaxis = list(title = "Lineage"),
    yaxis = list(title = "Counts"),
    legend = list(title = list(text = "Sublineage"))
  )
```

#### DRTypes (WHO)

```{r}
plot_ly(
  data = resistance_counts_df,
  x = ~DRT,
  y = ~n,
  color = ~lineage,
  type = 'bar'
) |>
  layout(
    barmode = 'stack',
    xaxis = list(title = "Drug Resistance Type (WHO)"),
    yaxis = list(title = "Counts"),
    legend = list(title = list(text = "Lineage"))
  )
```

#### Resistance (WHO)
```{r}
# Unique antibiotics and lineages
antibiotics <- unique(monthly_lineage_counts$Antibiotic)
lineages <- unique(monthly_lineage_counts$lineage)

# Prepare a list of traces (one for each Antibiotic + Main_lineage combo)
traces <- list()
visible <- rep(FALSE, length(antibiotics) * length(lineages))

idx <- 1
for (ab in antibiotics) {
  for (lin in lineages) {
    df_sub <- monthly_lineage_counts |>
      filter(Antibiotic == ab, lineage == lin)
    
    traces[[idx]] <- list(
      x = df_sub$month,
      y = df_sub$count,
      type = 'scatter',
      mode = 'lines+markers',
      name = lin,
      visible = FALSE
    )
    idx <- idx + 1
  }
}

# Initially show the first Antibiotic's traces
for (i in seq_along(lineages)) {
  visible[i] <- TRUE
}
for (i in seq_along(traces)) {
  traces[[i]]$visible <- visible[i]
}

# Create buttons for dropdown (one button per antibiotic)
buttons <- list()
for (i in seq_along(antibiotics)) {
  vis <- rep(FALSE, length(traces))
  start <- (i - 1) * length(lineages) + 1
  vis[start:(start + length(lineages) - 1)] <- TRUE
  
  buttons[[i]] <- list(
    method = "restyle",
    args = list("visible", vis),
    label = antibiotics[i]
  )
}

fig <- plot_ly()

for (tr in traces) {
  fig <- fig |> add_trace(x = tr$x, y = tr$y, type = tr$type, mode = tr$mode, name = tr$name, visible = tr$visible)
}

fig <- fig |> layout(
  title = "Monthly Resistant Counts by Main Lineage",
  xaxis = list(title = "Month"),
  yaxis = list(title = "Resistant Count"),
  updatemenus = list(
    list(
      buttons = buttons,
      direction = "down",
      x = 0,                   # left side of plot
      xanchor = "left",        # anchor the dropdown to the left
      y = 1,
      yanchor = "bottom",
      showactive = TRUE
    )
  )
)

fig
```


#### Failed QC

If this table is empty, then great! All your samples in the RutiSeq/db pass the quality metrics defined when you ran the RutiSeq workflow. Any failed variables will be colored in red.

```{r}
# Build datatable with conditional formatting
DT::datatable(
  failed_samples_df,
  options = list(pageLength = 10, autoWidth = TRUE),
  filter = 'top'
  ) |>
  formatStyle(
    'Total Reads',
    color = styleInterval(minReads, c('red', 'black'))
  ) |>
  formatStyle(
    'Unambiguous Total Bases (%)',
    color = styleInterval(minCovPerc, c('red', 'black'))
  ) |>
  formatStyle(
    'Unambiguous Coverage median',
    color = styleInterval(minCov, c('red', 'black'))
  ) |>
  formatStyle(
    'Infection type',
    color = styleEqual("Mixed", "red")
  )
```

#### SNP freq (main)
This tab features a ridge plot displaying multiple histograms that show the number of SNP variations identified between isolates belonging to each lineage and the reference genome (H37Rv). For clarity, isolate identified as a mixed infection have been removed. This is of value to understand the overall SNP frequencies of your dataset, and can be used to help identify samples that have suspicoiusly high or low number of SNPs (this can sometimes indicate there is something unusualy about the isolate - which can in turn cause issues with clustering)

```{r}
snp_freq <- analysisSummary_df |> 
    filter(`Infection type` != "Mixed") |>
    select(lineage = `Main lineage`, nSNPs)
ggplot(snp_freq, 
        aes(x = nSNPs, 
            y = lineage, 
            )
          ) +
        ggridges::geom_density_ridges(alpha=0.6, 
                                      #stat="binline", 
                                      bins=100) +
        ggridges::theme_ridges() + 
        theme(legend.position = "none") +
    labs(x = "No. SNPs", y = "Lineage")
```


#### SNP freq (sub)
This tab features a ridge plot displaying multiple histograms that show the number of SNP variations identified between isolates belonging to each sub-lineage and the reference genome (H37Rv). For clarity, isolate identified as a mixed infection have been removed. This is of value to understand the overall SNP frequencies of your dataset, and can be used to help identify samples that have suspicoiusly high or low number of SNPs (this can sometimes indicate there is something unusualy about the isolate - which can in turn cause issues with clustering)

```{r}
snp_freq <- analysisSummary_df |> 
  filter(`Infection type` != "Mixed") |>
  select(lineage = `Sub-lineage`, nSNPs) |>
  mutate(lineage = sapply(str_split(lineage, "\\."), function(parts) {
    if (length(parts) >= 2) {
      paste(parts[1], parts[2], sep = ".")
    } else {
      parts[1]
    }
  }))

ggplot(snp_freq, 
        aes(x = nSNPs, 
            y = lineage)) +
        ggridges::geom_density_ridges(alpha=0.6, 
                                      #stat="binline", 
                                      bins=100) +
        ggridges::theme_ridges() + 
        theme(legend.position = "none") +
    labs(x = "No. SNPs", y = "Sub-lineage")
```

:::

## Outbreak Clusters

```{r}
## Table of processedClusters_df# Reorder the full data frame
cluster_table <- processedClusters_df |>
  select(SampleID, lineage, all_of(t_cols_sorted))

## Clusters by lineage
cluster_table_long <- cluster_table |> 
  pivot_longer(cols = all_of(t_cols_sorted),
                names_to = "threshold",
                values_to = "clusterID")

cluster_table_long_counts <- cluster_table_long |> 
  select(lineage, threshold, clusterID) |>
  distinct() |>
  count(lineage, threshold)

# Example data frame (replace with your actual cluster data)
cluster_table_meta_df <- left_join(processedClusters_df, metadata_df, 
                                  by = c("SampleID" = "sampleID"))
# Sample data (replace this with your real data)
cluster_table_meta_timeseries <- 
    cluster_table_meta_df |>
    pivot_longer(
      cols = all_of(t_cols_sorted),
      names_to = "threshold",
      values_to = "clusterID"
    ) |>
    select(lineage, date, threshold, clusterID) |>
    mutate(date = as.Date(date)) |>
    filter(!is.na(clusterID)) |>
    mutate(month = floor_date(date, "month")) |>
    filter(!str_detect(clusterID, "^singleton-"))

# Calculate cumulative counts over time per clusterID and threshold
cumulative_counts_monthly <- cluster_table_meta_timeseries |>
  group_by(threshold, clusterID, lineage, month) |>
  summarise(count = n(), .groups = "drop") |>
  arrange(threshold, clusterID, month) |>
  group_by(threshold, clusterID, lineage) |>
  mutate(cumulative_count = cumsum(count)) |>
  ungroup()

## Newest clusters 
new_samples_clusterIDs <- cluster_table_long |> 
    filter(SampleID %in% new_samples$Sample) |> 
    select(clusterID)

old_samples_clusterIDs <- cluster_table_long |> 
    filter(!(SampleID %in% new_samples$Sample)) |> 
    select(clusterID)

# Select ONLY the new clusterIDs
newest_clusterIDs <- anti_join(new_samples_clusterIDs, old_samples_clusterIDs, by = "clusterID")
newest_clusters_df <- cluster_table_long |> filter(clusterID %in% newest_clusterIDs$clusterID) 
newest_clusters_df_wide <- newest_clusters_df |> 
    pivot_wider(
      names_from = threshold,
      values_from = clusterID
    )

# Select ONLY the growing clusters
growing_clusterIDs <- inner_join(new_samples_clusterIDs, old_samples_clusterIDs, by = "clusterID") |>
    filter(!str_detect(clusterID, "^singleton-"))
growing_clusters_df <- cluster_table_long |> filter(clusterID %in% growing_clusterIDs$clusterID) 
growing_clusters_df_wide <- growing_clusters_df |> 
    pivot_wider(
      names_from = threshold,
      values_from = clusterID
    )

cluster_table_long_new <- cluster_table_long |> 
    filter(clusterID %in% new_samples_clusterIDs$clusterID) |>
    filter(!str_detect(clusterID, "^singleton-"))
```

This section presents the outbreak clusters identified using SNP threshold clustering, based on the defined thresholds (`r paste(t_cols_sorted, collapse = ", ")`). Cluster identifiers follow the format: `Cluster number`**–**`SNP distance`**–**`Lineage`.

> ⚠️ Cluster identifiers are not preserved across analyses. For accurate interpretation, avoid comparing cluster IDs between different reports. If consistency is required, consider exporting the cluster assignments and creating your own internal naming convention, or updating the cluster files to remain consistent with past cluster identifies. This feature may change in the future, but there are currently no plans to implement persistent clusterIDs.

::: {.panel-tabset}

### Clusters

In this tab, a time series of clusters and their growth is displayed in the scatterplot. Use the slider to visualize the expansion of clusters over time. Each data point represents an outbreak cluster, colored by its lineage. For this plot, the lowest threshold is selected and only clusters larger than 3 cases.

```{r}
latest_cluster_counts <- cumulative_counts_monthly |>
  group_by(clusterID) |>
  slice_max(order_by = month, n = 1, with_ties = FALSE) |>
  ungroup()

# Extract and sort t= column names numerically
lowest_num <- min(as.numeric(gsub("t=", "", t_cols_sorted)))
lowest_threshold <- paste0("t=", lowest_num)

df_filtered <- latest_cluster_counts |>
  filter(threshold == lowest_threshold) |>
  filter(cumulative_count >= 3)

# Build the ggplot
p <- ggplot(
      data = df_filtered,
      aes(
        x = month,
        y = cumulative_count,
        size = cumulative_count,
        color = lineage,
        text = paste(
          "ClusterID:", clusterID,
          "<br>Lineage:", lineage,
          "<br>Threshold:", threshold,
          "<br>Count:", cumulative_count,
          "<br>Date:", month
        )
      )
  ) +
  geom_point(alpha = 0.7) +
  theme_minimal() +
  labs(
    x = "Last Update (most recent growth)",
    y = "Cluster size",
    color = "Lineage",
    size = ""
  )

# Convert to plotly
ggplotly(p, tooltip = "text")
```

### Growing clusters

In this tab, clusters identified within the RutiSeq-db that have grown with the addition of new cases are highlighted.

```{r}
DT::datatable(growing_clusters_df_wide, options = list(pageLength = 10, autoWidth = TRUE), filter = 'top')
```

### New clusters

In this tab, new clusters identified from the most recent analysis are presented. Each cluster is summarized by the number of isolates, resistance profiles, and the identities of the isolates within the cluster.

```{r}
DT::datatable(newest_clusters_df_wide, options = list(pageLength = 10, autoWidth = TRUE), filter = 'top')
```

:::

### Table of clusters

```{r}
# Render interactive table
DT::datatable(cluster_table, options = list(pageLength = 10, autoWidth = TRUE), filter = 'top')
```

---

# RutiSeq-db

Below is an interactive table of the entire RutiSeq-db for browsering. This includes all the antimicrobial resistance genes, resistanceprofiles, cluster memebership and general mapping statistics to the reference genome.

```{r}
# Render the datable
DT::datatable(analysisSummary_df_large, options = list(pageLength = 5, autoWidth = TRUE), filter = 'top')
```

---

# Methods

Reads were analysed using an custom workflow written with Nextflow, (RutiSeq-nf)[https://github.com/phesketh-igtp/RutiSeq-nf]. Raw FASTQ reads are taxonomically classified using Sylph v0.8.1 (microbial database, accessed 2025-06-02). Reads are analysed with *MTBseq v1.1.0* and *TB-Profiler v6.6.3*, which both perform referenced-based analysis using the virulent laboratory reference strain H37Rv (NC_000962) ​(Coll et al., 2014; Homolka et al., 2012; Kohl et al., 2018; Merker et al., 2015; Verboven et al., 2022)​. TB-Profiler classifies MTBC lineages and in silico drug-resistance profiling. TB-Profiler uses *Trimmomatic v0.39* ​(Bolger et al., 2014)​, BWA-mem v0.7.18 ​(Li & Durbin, 2009)​, and *freebayes v1.3.6* ​(Garrison & Marth, 2012)​ to perform adapter trimming, read mapping and variant calling respectively. Lineage classification was based on a 90-SNP barcode, while resistance profiles were derived using both WHO categorization (--db who), ​(The World Health Organisation, 2023)​. 

MTBseq conducted reference-based mapping for SNP calling, using *BWA v0.7.17*, *GATK v3.8* ​(McKenna et al., 2010)​, and *SAMtools v1.6* for mapping ​(Danecek et al., 2021)​, variant calling and mapping file managements respectively. We employed MTBSeq default values (i.e, four reads calling the allele, Phred score of 20, and minimum coverage of 75% allele frequency). Isolates with median coverage <50X and <90% were excluded from downstream analysis. SNP’s falling within annotated regions as ”phage”, ”maturase”, ”PE/PPE/PGRS”, or ”13E12 repeat family protein”, were excluded. 

Isolates are initially grouped by lineage (e.g. L1, L2, etc.). Lineage 4, the most abundant, was further subdivided into second-level sub-lineages (e.g. L4.1, L4.2, etc.) to reduce processing time. Lineage-grouped genomes are pairwise analysed using MTBseq functions (--step TBjoin, --step TBamend, and --step TBGroup), generating SNP distance matrices, and clustering at three distinct SNP distances (5, 10 and 15), and SNP multiple-sequence alignments for each (sub)-lineage. Maximum-likelihood trees (GTR+G4) were constructed from concatenated variable sites across lineage grouped genomes with *IQ-TREE v2.3.6*  ​(Minh et al., 2020; Nguyen et al., 2015)​. Resulting trees were visualized and annotated with the R package *ggtree v3.14* ​(Yu et al., 2017).​ Trees were manually inspected to identify any non-monophyletic clusters. For each genomic cluster, SNP were identified from MTBseq concatenated SNP alignment files using *SNP-sites v2.5.1* ​(Page et al., 2016)​. Reference position for cluster SNPs were extracted also with SNP-sites, and two reference SNP profiles were reconstructed, H37Rv and MTBc ancestral genomes reconstructed using ancient DNA ​(Comas et al., 2013)​. Resulting SNP alignments were converted to NEXUS format using seqret part of the Emboss v6.6.0 suit of tools ​(Rice et al., 2000)​.

## Parameters

The analysis parameters used in the workflow are summarized below, in addition to the default values.

| **Category**                      | Value                                                                 |
|-----------------------------------|-----------------------------------------------------------------------|
| **Run ID**                        | `r paste(runID)`                                                      |
| **Total cases**                   | `r nrow(analysisSummary_df)`                                          |
| **New cases**                     | `r nrow(new_samples)`                                                 |
| **SNP distance thresholds**       | `r paste(t_cols_sorted, collapse = ", ")` (*default: 5, 10, 15*)      |
| **Minimum Coverage (X)**          | `r paste(minCov, "X")` (*default: 50 X*)                              |
| **Minimum Coverage (%)**          | `r paste(100 * minCovPerc, "%")` (*default: 95 %*)                    |
| **Minimum Reads**                 | `r paste(format(minReads, big.mark = ",", scientific = FALSE))` (*default: 500,000*) |
| **IQ-TREE - no. bootstraps**      | `r paste(iqtree_bootstraps)`                                          |
| **IQ-TREE - model**               | `r paste(iqtree_model)`                                               |
| **MTBSeq - minbqual**             | `r paste(mtbseq_minbqual, "(_default: 13_)")`                         |
| **MTBSeq - mincovf**              | `r paste(mtbseq_mincovf, "(_default: 4_)")`                           |
| **MTBSeq - mincovr**              | `r paste(mtbseq_mincovr, "(_default: 4_)")`                           |
| **MTBSeq - minphred20**           | `r paste(mtbseq_minphred20, "(_default: 4_)")`                        |
| **MTBSeq - minfreq**              | `r paste(mtbseq_minfreq, "(_default: 75_)")`                          |
| **MTBSeq - unambig**              | `r paste(mtbseq_unambig, "(_default: 95_)")`                          |
| **MTBSeq - window**               | `r paste(mtbseq_window, "(_default: 12_)")`                           |

## References

​​Bolger, A. M., Lohse, M., & Usadel, B. (2014). *Trimmomatic: A Flexible Trimmer for Illumina Sequence Data*. Bioinformatics, 30(15), 2114–2120. [https://doi.org/10.1093/bioinformatics/btu170](https://doi.org/10.1093/bioinformatics/btu170)

​Coll, F., McNerney, R., Guerra-Assunção, J. A., Glynn, J. R., Perdigão, J., Viveiros, M., Portugal, I., Pain, A., Martin, N., & Clark, T. G. (2014). *A Robust SNP Barcode for Typing Mycobacterium Tuberculosis Complex Strains*. Nature Communications, 5(1), 4812. [https://doi.org/10.1038/ncomms5812](https://doi.org/10.1038/ncomms5812)

​Comas, I., Coscolla, M., Luo, T., Borrell, S., Holt, K. E., Kato-Maeda, M., Parkhill, J., Malla, B., Berg, S., Thwaites, G., Yeboah-Manu, D., Bothamley, G., Mei, J., Wei, L., Bentley, S., Harris, S. R., Niemann, S., Diel, R., Aseffa, A., … Gagneux, S. (2013). *Out-of-Africa Migration and Neolithic Coexpansion of Mycobacterium Tuberculosis with Modern Humans*. Nature Genetics, 45(10), 1176–1182. https://doi.org/10.1038/ng.2744 

​Danecek, P., Bonfield, J. K., Liddle, J., Marshall, J., Ohan, V., Pollard, M. O., Whitwham, A., Keane, T., McCarthy, S. A., Davies, R. M., & Li, H. (2021). *Twelve Years of SAMtools and BCFtools*. GigaScience, 10(2), giab008. [https://doi.org/10.1093/gigascience/giab008](https://doi.org/10.1093/gigascience/giab008)

​Garrison, E., & Marth, G. (2012). *Haplotype-based variant detection from short-read sequencing*. [https://arxiv.org/abs/1207.3907v2](https://arxiv.org/abs/1207.3907v2)

​Homolka, S., Projahn, M., Feuerriegel, S., Ubben, T., Diel, R., Nübel, U., & Niemann, S. (2012). *High Resolution Discrimination of Clinical Mycobacterium tuberculosis Complex Strains Based on Single Nucleotide Polymorphisms*. PLOS ONE, 7(7), e39855. [https://doi.org/10.1371/JOURNAL.PONE.0039855](https://doi.org/10.1371/JOURNAL.PONE.0039855)

​Kohl, T. A., Utpatel, C., Schleusener, V., Filippo, M. R. De, Beckert, P., Cirillo, D. M., & Niemann, S. (2018). *MTBseq: A Comprehensive Pipeline for Whole Genome Sequence Analysis of Mycobacterium Tuberculosis Complex Isolates*. PeerJ, 6, e5895. [https://doi.org/10.7717/peerj.5895](https://doi.org/10.7717/peerj.5895) 

​Li, H., & Durbin, R. (2009). *Fast and Accurate Short Read Alignment with Burrows-Wheeler Transform*. Bioinformatics (Oxford, England), 25(14), 1754–1760. [https://doi.org/10.1093/bioinformatics/btp324](https://doi.org/10.1093/bioinformatics/btp324) 

​McKenna, A., Hanna, M., Banks, E., Sivachenko, A., Cibulskis, K., Kernytsky, A., Garimella, K., Altshuler, D., Gabriel, S., Daly, M., & DePristo, M. A. (2010). *The Genome Analysis Toolkit: A MapReduce framework for analyzing next-generation DNA sequencing data*. Genome Research, 20(9), 1297–1303. [https://doi.org/10.1101/GR.107524.110](https://doi.org/10.1101/GR.107524.110) 

​Merker, M., Blin, C., Mona, S., Duforet-Frebourg, N., Lecher, S., Willery, E., Blum, M. G. B., Rüsch-Gerdes, S., Mokrousov, I., Aleksic, E., Allix-Béguec, C., Antierens, A., Augustynowicz-Kopeć, E., Ballif, M., Barletta, F., Beck, H. P., Barry, C. E., Bonnet, M., Borroni, E., … Wirth, T. (2015). *Evolutionary history and global spread of the Mycobacterium tuberculosis Beijing lineage*. Nature Genetics, 47(3), 242–249. [https://doi.org/10.1038/NG.3195](https://doi.org/10.1038/NG.3195)

​Minh, B. Q., Schmidt, H. A., Chernomor, O., Schrempf, D., Woodhams, M. D., von Haeseler, A., & Lanfear, R. (2020). *IQ-TREE 2: New Models and Efficient Methods for Phylogenetic Inference in the Genomic Era*. Molecular Biology and Evolution, 37(5), 1530–1534. [https://doi.org/10.1093/molbev/msaa015](https://doi.org/10.1093/molbev/msaa015) 

​Nguyen, L.-T., Schmidt, H. A., von Haeseler, A., & Minh, B. Q. (2015). *IQ-TREE: A Fast and Effective Stochastic Algorithm for Estimating Maximum-Likelihood Phylogenies*. Molecular Biology and Evolution, 32(1), 268–274. [https://doi.org/10.1093/molbev/msu300](https://doi.org/10.1093/molbev/msu300) 

​Page, A. J., Taylor, B., Delaney, A. J., Soares, J., Seemann, T., Keane, J. A., & Harris, S. R. (2016). *SNP-sites: Rapid Efficient Extraction of SNPs from Multi-FASTA Alignments*. Microbial Genomics, 2(4), e000056. [https://doi.org/10.1099/mgen.0.000056](https://doi.org/10.1099/mgen.0.000056)

​Rice, P., Longden, L., & Bleasby, A. (2000). *EMBOSS: the European Molecular Biology Open Software Suite*. Trends in Genetics : TIG, 16(6), 276–277. [https://doi.org/10.1016/S0168-9525(00)02024-2](https://doi.org/10.1016/S0168-9525(00)02024-2) 

​The World Health Organisation. (2023). *Catalogue of Mutations in Mycobacterium Tuberculosis Complex and Their Association with Drug Resistance  : Second Edition*. World Health Organization. 

​Verboven, L., Phelan, J., Heupink, T. H., & Rie, A. Van. (2022). *TBProfiler for Automated Calling of the Association with Drug Resistance of Variants in Mycobacterium Tuberculosis*. PLOS ONE, 17(12), e0279644. [https://doi.org/10.1371/journal.pone.0279644](https://doi.org/10.1371/journal.pone.0279644)

​Yu, G., Smith, D. K., Zhu, H., Guan, Y., & Lam, T. T.-Y. (2017). *ggtree: An r Package for Visualization and Annotation of Phylogenetic Trees with Their Covariates and Other Associated Data*. Methods in Ecology and Evolution, 8(1), 28–36. [https://doi.org/10.1111/2041-210X.12628](https://doi.org/10.1111/2041-210X.12628)