---
title: "RutiSeq Report"
format: html
execute:
    echo: false
    warning: false
    message: false
---

# RutiSeq Report — Run ID: `r run_id`

```{r setup, include=FALSE}
# Load required libraries
library(readr)
library(dplyr)
library(ggplot2)
library(plotly)
library(DT)

# Read environment variables
run_id         <- Sys.getenv("runID")
summary_file   <- Sys.getenv("analysisSummary")
cluster_file   <- Sys.getenv("processedClusters")
who_file       <- Sys.getenv("whoResistance")
tbdb_file      <- Sys.getenv("tbdbResistance")
report_dir     <- Sys.getenv("reportDir")

# Load datasets
summary_data   <- read_tsv(summary_file)
clusters_data  <- read_tsv(cluster_file)
who_data       <- read_tsv(who_file)
tbdb_data      <- read_tsv(tbdb_file)
```

---

# RutiSeq Report — Run ID: `r run_id`

Cases - **Number of cases in `r run_id`:** `r run_id` |  **Total cases in RutiSeq-db:** `r run_id`
Dates - **Most recent case:** `r run_id` | **Oldest case:** `r run_id`

---

## Database Summary

::: {.panel-tabset}

### `r run_id`

```{r}
# This table shows results only for the current run
datatable(summary_data %>% filter(run_id == run_id), 
            caption = paste("Results for run:", run_id))
```

### `RutiSeq-db`
```{r}
# This table shows results only for the current run
datatable(summary_data %>% filter(run_id == run_id), 
            caption = paste("Results for entire RutiSeq database))
```

:::

---

## MTBC Lineages

```{r}

```


::: {.panel-tabset}

### L1

```{r}

```

### L2

```{r}

```

### L3

```{r}

```

### L4

```{r}

```

### L5

```{r}

```

### L6
```{r}

```

:::



---

## Phylogeny

::: {.panel-tabset}

### L1

```{r}
rdata <- load(paste0(report_dir,"//lineage1.contree.Rdata")

```

### L2

```{r}

```

### L3

```{r}

```

### L4

```{r}

```

### L5

```{r}

```

### L6
```{r}

```

:::

---

## Clusters

> ⚠️ Note: Cluster identities are not preserved across runs. For accurate interpretation, avoid comparing cluster IDs between different reports.



---

## Cluster Growth